Multisequence algorithm for coarse-grained biomolecular simulations: exploring the sequence-structure relationship of proteins

### 2017-07-18 Manuscript.tex
Revision for JCP

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\section{Introduction}
\noindent
Recent years have seen important advances in biomolecular simulation methods, including improvements to standard molecular dynamics force fields,~\cite{Piana2014} the advent of several alternative atomistic simulation approaches,~\cite{Ding2008,Irback2006,Verma2009,Yang2007} and new techniques for  conformational sampling.~\cite{Bernardi2015} Together with the ever-increasing availability of computational resources, these advances have triggered a few major efforts~\cite{McGuffee2010,Miao2010,Perilla2016,Lindorff-Larsen2011,Yu2016} to characterize the dynamics of biomolecular systems of various sizes, e.g., a small native protein on the millisecond scale~\cite{Lindorff-Larsen2011} and a comprehensive model cytoplasm on the nanosecond scale.~\cite{Yu2016} While encouraging and insightful, these large-scale simulations have also highlighted the fact that severe tradeoffs in size and time scales will likely persist for the foreseeable future.

One way to expand the range of biomolecular simulations is to turn to coarse-grained (CG) models, where the basic aim is to simplify the physical description of interactions while retaining the essential physics of the system under study.~\cite{Riniker2012} Ingolfsson \textit{et al.} list 4 main factors that make CG models computationally fast: reduction in the number of degrees of freedom, faster simulation dynamics, emphasis on short-range interactions and the ability of using larger integration time steps.~\cite{Ingolfsson2014} To this list can be added that a CG representation of either the interaction potential or the molecular geometry often opens up for alternative sampling schemes beyond traditional molecular dynamics approaches, which can further speed up conformational sampling. Examples of such sampling schemes include activation-relaxation kinetics,~\cite{Beland2011} discrete molecular dynamics~\cite{Proctor2011} and various Monte Carlo (MC)-based techniques such as cluster moves.~\cite{Vitalis2009}

The challenges of achieving representative conformational sampling of individual biomolecular systems notwithstanding, many biologically motivated problems naturally call for the investigation and comparison of molecular variants, e.g., determining the molecular mechanisms of specificity in protein-protein~\cite{Zarrinpar2003,Hakes2007} and protein-nucleotide~\cite{Rohs2010} interactions, or the role of mutations in molecular disease processes.~\cite{Ross2004} Another example is protein folding, where unique insight has been achieved by comparing sequences within and between protein families.~\cite{Tzul2017,Wensley2010} In a situation with extremely rapid growth of sequence information,~\cite{Vukmirovic2000} it is of interest to explore ways to efficiently sample multiple sequences in biomolecular simulations.

@@

%While this may be hard to achieve for detailed biomolecular models, it is likely possible in many CG models.

As an initial test case, we apply the MS method to a set of 144 model protein sequences with 16 amino acids. This set (denoted here {\SR}) was constructed to sparsely cover the sequence space between two ideally designed sequences, A1 and N1, which fold into an $\alpha$-helix and a $\beta$-hairpin,~\cite{Bhattacherjee2012} respectively, as shown in Fig.~2. In a previous work, this set allowed us to demonstrate the existence of A1-N1 mutational pathways which do not proceed through any disordered (non-folding) sequences but instead switch abruptly between the two folds in a single mutational step.~\cite{Holzgrafe2014} Such an ability to reversibly switch folds was recently demonstrated in a handful of natural and engineered proteins.~\cite{Bryan2010} Here we use the set {\SR} and our CG model for protein folding to validate the MS method and compare its efficiency to a standard generalized-ensemble method.~\cite{Marinari1992,Lyubartsev1992} We thereafter greatly enlarge {\SR} to a set with 1,024 sequences, also spanning the A1-N1 space, as well as another set of the same size spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively (see Fig.~2). Besides demonstrating that the MS method can be applied to large numbers of sequences, the results allow us to carry out a more systematic analysis of the biophysical properties of sequences along mutational pathways connecting these two pairs of basic folds than has been previously possible.

\begin{figure}
\includegraphics[width=4.2cm]{Fig1}
\caption{The two types of Monte Carlo updates in the multisequence Monte Carlo algorithm.}
\end{figure}

@@

\subsection{Generalized-ensemble algorithms and simulated tempering}
\noindent
Conventional Monte Carlo simulations of the canonical distribution is problematic at low temperatures $T$ for many physical systems because simulations tend to become trapped in local energy minima and hamper representative sampling of configurational space. The basic idea of generalized-ensemble algorithms~\cite{Mitsutake2001} is to alleviate this trapping problem by sampling states using a non-Boltzmann weight factor and/or expand the state space with additional dynamical parameters.

@@

\subsection{Multisequence algorithm}
\noindent 
The basic idea of the MS algorithm for biomolecular simulation is to let the sequence $\sbar$ become a dynamic parameter rather than the temperature as in ST. A dynamic $\sbar$ is technically feasible when the potential energy can be written as $E(\sbar,\rbar)$, where $\sbar$ and $\rbar$ are independent variables. This is the case in our coarse-grained protein model which has only backbone degrees of freedom. It can also be achieved for in some more detailed models.~\cite{Bhattacherjee2013,Wallin2017}

Similarly to ST, the MS algorithm simulates the joint probability distribution
\begin{equation}
P(\sbar,\rbar) =\dfrac{1}{Z}e^{-\beta E(\sbar,\rbar) + h(\sbar)}\,, 
\label{MS}
\end{equation}
where  
\begin{equation}
Z = \sum_{\sbar}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)}\,
\end{equation}
and the first sum goes over a set of allowed sequences $\sbar$. The simulation parameters $h(\sbar)$, similar to the parameters $g_m$ in ST, control the marginal distribution $P(\sbar)=Z^{-1}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)} = \tilde{Z}^{-1}e^{-\beta F(\sbar)+ h(\sbar)}$ and a roughly flat $P(\sbar)$ can be achieved by choosing $h(\sbar) \approx \beta F(\sbar)$, where $F(\sbar)$ is the free energy of sequence $\sbar$ at temperature $T$.

Two types of MC updates are required to sample from the distrubution in Eq.~\ref{MS}, ordinary conformational update $\rbar\rightarrow\rbar'$ and mutational updates $\sbar\rightarrow\sbar'$. The acceptance probabilities for the latter  becomes
\begin{equation}
P_\mathrm{acc} (\sbar\rightarrow\sbar') = \min [1, \exp\{-\beta\Delta E+\Delta h\}]\,,
\label{accrej}
\end{equation}
where $\Delta E = E(\sbar',\rbar)-E(\sbar,\rbar)$ and $\Delta h = h({\sbar'})-h(\sbar)$.

\section{Model and Methods}
\subsection{Coarse-grained 3-letter model for protein folding}
\noindent
All calculations were carried out using the coarse-grained model for protein folding developed in Ref.~\citenum{Bhattacherjee2012}. In this model, there are 3 different amino acid types: hydrophobic (h), polar (p) and turn-type (t). The backbone chain is represented atomistically by the N, H, $\mathrm{C}_\alpha$, $\mathrm{H}_{\alpha 1}$, C$'$ and O atoms. By contrast, the sidechain represention is simplified to a single enlarged $\mathrm{C}_\beta$ atom, which is geometrically identical for h and p types. The sidechain is absent for the t type which instead has an $\mathrm{H}_{\alpha 2}$ atom. The t type is therefore closely related to glycine. All bond lengths, bond angles, and peptide plane angles (180$^\circ$) are held fixed. Hence, an $N$-amino acid chain conformation $\rbar$ can, for any sequence $\sbar$, therefore be described by the set of 2$N$ backbone torsional angles $\{\phi_i$, $\psi_i\}_{i=1}^{N}$.

@@

\subsection{Model sequences}
\noindent
Six of the model sequences studied in this work, A1, N1, R1, R2, A2, and TN, are given in Table~I. In addition, we study two sequence sets $\mathrm{S16}_{1024}$  and $\mathrm{S35}_{1024}$ with 1,024 sequences each derived from the A1-N1 and A2-TN pairs, respectively, through mutational combinations, as well as the set $\mathrm{S16}_{144}$ taken from  Ref.~\protect\citenum{Holzgrafe2014}.

@@

\subsection{Monte Carlo simulation parameters}
\noindent
Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assign a new value between $-\pi$ and $\pi$; and (2) a semi-local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2001} In MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are carried out by randomly picking a new sequence $\sbar'\ne\sbar$ and applying the accept-reject criterion in Eq.~\ref{accrej}. A sequence move is attempted every $1,000$ MC steps while temperature updates $m\rightarrow m'$ are attempted every 100 steps. The simulations carried out in this work are summarized in Table~II.

@@

\subsection{Observables}
\noindent
Fold stabilities are calculated as in Ref.~\citenum{Holzgrafe2015} and described briefly below. First we define two structural similarity measures $\QIA$ and $\QIB$ for folds IA and IB, respectively, indicating the fraction of the fold-specific backbone-backbone hydrogen bonds that have been formed. The fold IA-hydrogen bonds are (2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15) and the fold IB-bonds are (3,14), (5,12), (7,10), (10,7), (12,5), (14,3), where (i,j) indicates a hydrogen bond between the CO group of amino acid i and the NH group of amino acid j. The stabilities of folds IA and IB are defined as the probabilities $\PIA = P(\QIA\ge0.8)$ and $\PIB = P(\QIB\ge0.8)$, respectively, i.e., the probability that at least 80\% of the fold's hydrogen bonds are formed. $\PIA$ and $\PIB$ thus depend on both sequence $\sbar$ and temperature $T$. For example, $\PIA=0.875\pm 0.003$ for A1 and $\PIB=0.785\pm0.008$ for N1 at $\kb T = 0.43$. Structural similarity measures for 35-amino acid folds IIA and IIB are defined as $\QIIA = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IA}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$ and $\QIIB = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IB}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$, respectively, where superscripts on $\QIA$ and $\QIB$ indicate over which amino acid positions those measures are applied to within the longer 35 amino acid sequences and $Q_\mathrm{tert}$ is a measure that counts the number of $\mathrm{C}_\beta$-$\mathrm{C}_\beta$ contacts between the two secondary structure elements of these folds.~\cite{Holzgrafe2015} In analogy with $\PIA$ and $\PIB$, we define the stabilities of folds IIA and IIB as $\PIIA = P(\QIIA\ge0.8)$ and $\PIIB = P(\QIIB\ge0.8)$, respectively. The root-mean-square-deviation, RMSD, is calculated over all $\mathrm{C}_\alpha$ atoms.

\begin{figure}
\includegraphics[width=8.3cm]{Fig2}
\caption{(A) Example of an MS simulation applied to the sequence set $\mathrm{S16}_{144}$. The plot shows the MC evolution of the sequence $\sbar$ (numbered 1--144), the total potential energy $E$ and the root-mean-square deviation (RMSD) calculated against fold IA (light blue) and fold IB (dark red) structures shown in (B). The simulation is carried out at $\kb T = 0.43$. (B) Representative structures of fold IA and fold IB, selected to be the minimum-energy conformations found for the sequences A1 and N1, respectively. (C) Representative fold IIA and fold IIB structures selected to be the minimum-energy conformations of A2 and TN, respectively. }
\end{figure}

@@

\subsection{Computational efficiency}
\noindent
We start by applying the MS algorithm to the set {\SR} across a range of temperatures $T$ (see Table~II). Two of the sequences in {\SR} are A1 and N1 (see Table~I) which fold into stable $\alpha$-helix and $\beta$-hairpin structures, respectively, as shown in Fig.~2B. Because A1 and N1 differ at 10 positions, 10 consecutive point mutations can transform A1 into N1, and vice versa. The binary sequence space between A1 and N1 in which any combination of these mutations have been carried out, therefore contains $2^{10}=1,024$ sequences. The 144 sequences in {\SR} were selected from this binary space with the constraints that the total number of hydrophobic amino acids are not too high and that they are not too unevenly distributed along the sequence.~\cite{Holzgrafe2014}

@@

One might have suspected that the MS algorithm would be hampered by poor acceptance rates for sequence updates. However, this is not the case in our model. We carry out updates $\sbar\rightarrow\sbar'$ by picking a new random sequence $\sbar'\ne\sbar$ from the set of allowed sequences. The (average) acceptance rate  depends on both $T$ and the step in sequences space $\Delta h$, i.e., the number of amino acid positions changed, as shown in Fig.~3. At the lowest $T$ and highest $\Delta h$, acceptance rates are only around 0.1-0.2. However, for most other $T$ and $\Delta h$ the overall acceptance rate is substantially higher and often above the oft-quoted rule-of-thumb value 0.25~\cite{Gilks1996} (see Fig~4B). An increased acceptance rate can easily be achieved by restricting proposed updates such that $\Delta h\le\Delta h_\mathrm{max}$, where $\Delta h_\mathrm{max}$ is a maximum step size, which might be necessary for longer chains.

\begin{figure}
\includegraphics[width=8.0cm]{Fig4}
\caption{Comparing sampling efficiency of the MS and ST algorithms. Statistical errors $\sigma_{\left < E\right >}$ of the average total energy $\left < E\right >$ from the two methods for the 16-amino acid sequences (A) A1, (B) N1, (C) R1 and (D) R2. Simulation lengths in the two methods are adjusted such that the number of conformations sampled per sequence and temperature is roughly the same (see text). }
\end{figure}

@@

\begin{figure*}
\rotatebox{0}{\includegraphics[width=16.6cm]{Fig5}}
\caption{Networks of sequences connecting folds IA and IB (top) and folds IIA and IIB (bottom). Each node represents a stable sequence ($\Ptot\ge\Pcut$ where $\Pcut=0.50$) that folds into either IA or IIA (light blue), IB or IIB (dark red), or is classified as bistable ($B>0.5$, black). A line between two nodes indicates that the sequences differ at only one position. Graph created using the tool Graphviz~\protect\cite{Graphviz2000} obtained from  www.graphviz.org.}
\end{figure*}

\vspace{12pt}
\subsection{Exploring sequence space: IA/IB and IIA/IIB fold connectivities}
\noindent
We now turn to the full binary sequence sets {\SI} and {\SII} with 1,024 sequences each. By applying the MS method to these two sets (see Table~II), we determine the low-$T$ thermodynamic behavior of each included sequence. In particular, we calculate the stabilities of folds IA and IB, $\PIA$ and $\PIB$, for all sequences in {\SI} and the stabilities of folds IIA and IIB, $\PIIA$ and $\PIIB$, for all sequences in {\SII} (see Methods). The relative statistical errors on these quantities vary but are only a few percent at the most, despite the large number of sequences included.

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\section{Introduction}

Recent years have seen important advances in biomolecular simulation methods, including improvements to standard molecular dynamics force fields,~\cite{Piana2014} the advent of several alternative atomistic simulation approaches,~\cite{Ding2008,Irback2006,Verma2009,Yang2007} and new techniques for  conformational sampling.~\cite{Bernardi2015} Together with the ever-increasing availability of computational resources, these advances have triggered a few major efforts~\cite{McGuffee2010,Miao2010,Perilla2016,Lindorff-Larsen2011,Yu2016} to characterize the dynamics of biomolecular systems of various sizes, e.g., a small native protein on the millisecond scale~\cite{Lindorff-Larsen2011} and a comprehensive model cytoplasm on the nanosecond scale.~\cite{Yu2016} While encouraging and insightful, these large-scale simulations have also highlighted that severe tradeoffs in size and time scales will likely persist for the foreseeable future.

One way to expand the range of biomolecular simulations is to turn to coarse-grained (CG) models, where the basic aim is to simplify the description of physical interactions while retaining the essential physics of the system under study.~\cite{Riniker2012} Ingolfsson \textit{et al.} list 4 main factors that make CG models computationally fast: reduction in the number of degrees of freedom, faster simulation dynamics, emphasis on short-range interactions and the ability of using larger integration time steps.~\cite{Ingolfsson2014} To this list can be added that a CG representation of either the interaction potential or the molecular geometry often opens up for alternative sampling schemes beyond traditional molecular dynamics approaches, which can further speed up conformational sampling. Examples of such sampling schemes include activation-relaxation kinetics,~\cite{Beland2011} discrete molecular dynamics~\cite{Proctor2011} and various Monte Carlo (MC)-based techniques such as cluster moves.~\cite{Vitalis2009}

The challenges of achieving representative conformational sampling of individual biomolecular systems notwithstanding, many biologically motivated problems naturally call for the investigation and comparison of molecular variants, e.g., determining the molecular mechanisms of specificity in protein-protein~\cite{Zarrinpar2003,Hakes2007} or protein-nucleotide~\cite{Rohs2010} interactions, or the role of mutations in molecular disease processes.~\cite{Ross2004} Another example is protein folding, where unique insight has been achieved by comparing sequences within and between protein families.~\cite{Tzul2017,Wensley2010} In a situation with extremely rapid growth of sequence information,~\cite{Vukmirovic2000} it is of interest to explore ways to efficiently sample multiple sequences in biomolecular simulations.

@@

{\color{red} As a test case for the MS algorithm, we selected the phenomenon of protein fold switching~\cite{Bryan2010} which has been recently demonstrated in a handful of natural and engineered proteins. These special proteins exhibit a unique ability to reversibly switch between different folds with accompanying changes in secondary structure, hydrophobic core packing and overall shape. Fold switching in proteins often plays a functional role. For example, rare transitions to an alternative fold in the protein KaiB provide a crucial time delay mechanism in the circadian clock cycle of cyanobacteria.~\cite{LiWang2015} Fold switching can be triggered by various events including changes to solution conditions,~\cite{Kuloglu2002} subdomain detachment,~\cite{Burmann2012} ligand binding~\cite{Luo2004} and point mutations.~\cite{Alexander2009} }

{\color{red} The discovery that abrupt and major changes in the 3-dimensional structures of proteins can be driven by the accumulation of point mutations, in particular, holds implications for protein evolution as it suggests a simple mechanism of fold evolution. The A and B domains of protein G fold into 3$\alpha$ and 4$\beta+\alpha$ structures, respectively,~\cite{Alexander2009} and are thus unrelated in both structure and sequence. Yet Alexander \textit{et al.}~\cite{Alexander2009} demonstrated that, after extensive initial mutations to both domains that left their respective folds undisturbed, a single additional mutation triggered an abrupt fold switch. This possibility suggests a potential evolutionary relationship between these two domains despite the lack of detectable similarity in both sequence or structure in wild-type protein G.~\cite{He2012} It is still unclear, however, how common such fold-to-fold mutational pathways are in evolutionary processes. We showed previously~\cite{Holzgrafe2014,Holzgrafe2015} that, within the model framework provided by our CG model,~\cite{Bhattacherjee2012} there are mutational paths between elementary protein folds that do not proceed through any disordered (non-folding) sequences. Instead, these paths exhibit abruptly switching between folds in a single, or very few, mutational steps similarly to the A and B domains of protein G. }

In demonstrating this fold-switching behavior in our model we characterized the folding of a set of 144 different model sequences with 16 amino acids. This set (denoted here {\SR}) was constructed to sparsely span the sequence space between two ideally designed sequences, A1 and N1, folding into an $\alpha$-helix and a $\beta$-hairpin, respectively, as shown in Fig.~2. Here we use the set {\SR} to validate the MS method and compare its computational efficiency to a standard generalized-ensemble method.~\cite{Marinari1992,Lyubartsev1992} We thereafter greatly enlarge {\SR} to a set with 1,024 sequences as well as another set of the same size spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively (see Fig.~2). Besides demonstrating that the MS method can be applied to large numbers of sequences, the results allow us to carry out a more systematic analysis of the biophysical properties of sequences along mutational pathways connecting these two pairs of basic folds than has been previously possible.

%As an initial test case, we apply the MS method to a set of 144 model protein sequences with 16 amino acids. This set (denoted here {\SR}) was constructed to sparsely cover the sequence space between two ideally designed sequences, A1 and N1, which fold into an $\alpha$-helix and a $\beta$-hairpin,~\cite{Bhattacherjee2012} respectively, as shown in Fig.~2. In a previous work, this set allowed us to demonstrate the existence of A1-N1 mutational pathways which do not proceed through any disordered (non-folding) sequences but instead switch abruptly between the two folds in a single mutational step.~\cite{Holzgrafe2014} Such an ability to reversibly switch folds was recently demonstrated in a handful of natural and engineered proteins.~\cite{Bryan2010} Here we use the set {\SR} and our CG model for protein folding to validate the MS method and compare its efficiency to a standard generalized-ensemble method.~\cite{Marinari1992,Lyubartsev1992} We thereafter greatly enlarge {\SR} to a set with 1,024 sequences, also spanning the A1-N1 space, as well as another set of the same size spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively (see Fig.~2). Besides demonstrating that the MS method can be applied to large numbers of sequences, the results allow us to carry out a more systematic analysis of the biophysical properties of sequences along mutational pathways connecting these two pairs of basic folds than has been previously possible.

\begin{figure}
\includegraphics[width=4.2cm]{./Fig1}
\caption{The two types of Monte Carlo updates in the multisequence Monte Carlo algorithm.}
\end{figure}

@@

\subsection{Generalized-ensemble algorithms and simulated tempering}

Conventional Monte Carlo simulations of the canonical distribution is problematic at low temperatures $T$ for many physical systems because simulations tend to become trapped in local energy minima and hamper representative sampling of configurational space. The basic idea of generalized-ensemble algorithms~\cite{Mitsutake2001} is to alleviate this trapping problem by sampling states using a non-Boltzmann weight factor and/or expand the state space with additional dynamical parameters.

@@

\subsection{Multisequence algorithm}

The basic idea of the MS algorithm for biomolecular simulation is to let the sequence $\sbar$ become a dynamic parameter rather than the temperature as in ST. A dynamic $\sbar$ is technically feasible when the potential energy can be written as $E(\sbar,\rbar)$, where $\sbar$ and $\rbar$ are independent variables. This is the case in our coarse-grained protein model which has only backbone degrees of freedom. It can also be achieved in some more detailed models.~\cite{Bhattacherjee2013,Wallin2017}

Similarly to ST, the MS algorithm simulates the joint probability distribution
\begin{equation}
P(\sbar,\rbar) =\dfrac{1}{Z}e^{-\beta E(\sbar,\rbar) + h(\sbar)}\,, 
\label{MS}
\end{equation}
where  {\color{red}
\begin{equation}
Z = \sum_{\sbar\in\mathrm{S}}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)}\,
\label{eq:ZMS}
\end{equation}
and S is a set of pre-selected sequences, i.e., the sequences to which visits are allowed during a simulation.} The simulation parameters $h(\sbar)$, similar to the parameters $g_m$ in ST, control the marginal distribution $P(\sbar)=Z^{-1}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)} = \tilde{Z}^{-1}e^{-\beta F(\sbar)+ h(\sbar)}$ and a roughly flat $P(\sbar)$ can be achieved by choosing $h(\sbar) \approx \beta F(\sbar)$, where $F(\sbar)$ is the free energy of sequence $\sbar$ at temperature $T$.

Two types of MC updates are required to sample from the distrubution in Eq.~\ref{MS}, ordinary conformational updates $\rbar\rightarrow\rbar'$ and {\color{red} sequence} updates $\sbar\rightarrow\sbar'$. The acceptance probabilities for the latter  becomes
\begin{equation}
P_\mathrm{acc} (\sbar\rightarrow\sbar') = \min [1, \exp\{-\beta\Delta E+\Delta h\}]\,,
\label{accrej}
\end{equation}
where $\Delta E = E(\sbar',\rbar)-E(\sbar,\rbar)$ and $\Delta h = h({\sbar'})-h(\sbar)$.

\section{Model and Methods}
\subsection{Coarse-grained 3-letter model for protein folding}

All calculations were carried out using the coarse-grained model for protein folding developed in Ref.~\citenum{Bhattacherjee2012}. In this model, there are 3 different amino acid types: hydrophobic (h), polar (p) and turn-type (t). The backbone chain is represented atomistically by the N, H, $\mathrm{C}_\alpha$, $\mathrm{H}_{\alpha 1}$, C$'$ and O atoms. By contrast, the sidechain represention is simplified to a single enlarged $\mathrm{C}_\beta$ atom, which is geometrically identical for h and p types. The sidechain is absent for the t type which instead has an $\mathrm{H}_{\alpha 2}$ atom. The t type is therefore closely related to glycine. All bond lengths, bond angles, and peptide plane angles (180$^\circ$) are held fixed. Hence, an $N$-amino acid chain conformation $\rbar$ can, for any sequence $\sbar$, therefore be described by the set of 2$N$ backbone torsional angles $\{\phi_i$, $\psi_i\}_{i=1}^{N}$.

@@

\subsection{Model sequences}

Six of the model sequences studied in this work, A1, N1, R1, R2, A2, and TN, are given in Table~I. In addition, we study two sequence sets $\mathrm{S16}_{1024}$  and $\mathrm{S35}_{1024}$ with 1,024 sequences each derived from the A1-N1 and A2-TN pairs, respectively, through mutational combinations, as well as the set $\mathrm{S16}_{144}$ taken from  Ref.~\protect\citenum{Holzgrafe2014}.

@@

\subsection{Monte Carlo simulation parameters and updates}

Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assign a new value between $-\pi$ and $\pi$; and (2) a semi-local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2001}

%, and where S is the set of allowed (pre-selected) sequences, S (see Eq.~\ref{eq:ZMS})
%i.e., given by the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$ angles,

{\color{red}In MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are carried out in the following way. First, a new random sequence $\sbar$ is picked from the set of pre-selected (allowed) sequences S, such that $\sbar'\ne\sbar$. The new sequence $\sbar'$ therefore differs from $\sbar$ in one or more amino acid positions. Thereafter, the sidechains of the protein, which remains in an unchanged (backbone) conformation $\rbar$, is re-built according to the new sequence $\sbar'$. Practically this means that, at the position(s) where the amino acid type has changed, the sidechain is altered according to the type change. For example, if $\mathrm{p}\rightarrow\mathrm{t}$, the $\mathrm{C}_\beta$ atom is removed and replaced with an $\mathrm{H}_{\alpha 2}$ atom or, if $\mathrm{p}\rightarrow\mathrm{h}$, the $\mathrm{C}_\beta$ remains in place but its character is changed to hydrophobic. Finally, the change in total energy $\Delta E$ is calculated and the accept-reject criterion in Eq.~\ref{accrej} is applied. If rejected, the old state $(\sbar,\rbar)$ is restored.}

A sequence update is attempted every $1,000$ MC steps while temperature updates $m\rightarrow m'$ are attempted every 100 steps. {\color{red}The computational cost for sequence updates is somewhat higher than for temperature updates, which does not require any energy calculations and is thus extremely rapid. For the purpose of comparing computational efficiencies between ST and MS, we therefore chose sequence updates to be slightly less frequent than temperature updates.} All simulations carried out in this work are summarized in Table~II.

@@

\subsection{Observables}

Fold stabilities are calculated as in Ref.~\citenum{Holzgrafe2015} and described briefly below. First we define two structural similarity measures $\QIA$ and $\QIB$ for folds IA and IB, respectively, indicating the fraction of the fold-specific backbone-backbone hydrogen bonds that have been formed. The fold IA-hydrogen bonds are (2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15) and the fold IB-bonds are (3,14), (5,12), (7,10), (10,7), (12,5), (14,3), where (i,j) indicates a hydrogen bond between the CO group of amino acid i and the NH group of amino acid j. The stabilities of folds IA and IB are defined as the probabilities $\PIA = P(\QIA\ge0.8)$ and $\PIB = P(\QIB\ge0.8)$, respectively, i.e., the probability that at least 80\% of the fold's hydrogen bonds are formed. $\PIA$ and $\PIB$ thus depend on both sequence $\sbar$ and temperature $T$. For example, $\PIA=0.875\pm 0.003$ for A1 and $\PIB=0.785\pm0.008$ for N1 at $\kb T = 0.43$. Structural similarity measures for 35-amino acid folds IIA and IIB are defined as $\QIIA = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IA}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$ and $\QIIB = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IB}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$, respectively, where superscripts on $\QIA$ and $\QIB$ indicate over which amino acid positions those measures are applied to within the longer 35 amino acid sequences and $Q_\mathrm{tert}$ is a measure that counts the number of $\mathrm{C}_\beta$-$\mathrm{C}_\beta$ contacts between the two secondary structure elements of these folds.~\cite{Holzgrafe2015} In analogy with $\PIA$ and $\PIB$, we define the stabilities of folds IIA and IIB as $\PIIA = P(\QIIA\ge0.8)$ and $\PIIB = P(\QIIB\ge0.8)$, respectively. The root-mean-square-deviation, RMSD, is calculated over all $\mathrm{C}_\alpha$ atoms.

\begin{figure}
\includegraphics[width=8.5cm]{Fig2}
\caption{(A) Example of an MS simulation of the sequence set $\mathrm{S16}_{144}$ carried out at $\kb T = 0.43$. The plot shows the MC evolution of the sequence $\sbar$ (numbered 1--144), the total potential energy $E$ and the root-mean-square deviation (RMSD) calculated against the representative fold IA (light blue) and fold IB (dark red) structures in (B). Representative structures of folds (B) IA, IB, (C) IIA and IIB, chosen to be the minimum-energy conformations found for the sequences A1, N1, A2 and TN, respectively. }
\end{figure}

@@

\subsection{Computational efficiency}

We start by applying the MS algorithm to the set {\SR} across a range of temperatures $T$ (see Table~II). Two of the sequences in {\SR} are A1 and N1 (see Table~I) which fold into stable $\alpha$-helix and $\beta$-hairpin structures, respectively, as shown in Fig.~2B. Because A1 and N1 differ at 10 positions, 10 consecutive point mutations can transform A1 into N1, and vice versa. The binary sequence space between A1 and N1 in which any combination of these mutations have been carried out, therefore contains $2^{10}=1,024$ sequences. The 144 sequences in {\SR} were selected from this binary space with the constraints that the total number of hydrophobic amino acids are not too high and that they are not too unevenly distributed along the sequence.~\cite{Holzgrafe2014}

@@

One might have suspected that the MS algorithm would be hampered by poor acceptance rates for sequence updates. However, this is not the case in our model. We carry out updates $\sbar\rightarrow\sbar'$ by picking a new random sequence $\sbar'\ne\sbar$ from the set of allowed sequences. The (average) acceptance rate  depends on both $T$ and the step in sequences space $\Delta h$, i.e., the number of amino acid positions changed, as shown in Fig.~3. At the lowest $T$ and highest $\Delta h$, acceptance rates are only around 0.1-0.2. However, for most other $T$ and $\Delta h$ the overall acceptance rate is substantially higher and often above the oft-quoted rule-of-thumb value 0.25~\cite{Gilks1996} (see Fig~4B). An increased acceptance rate can easily be achieved by restricting proposed updates such that $\Delta h\le\Delta h_\mathrm{max}$, where $\Delta h_\mathrm{max}$ is a maximum step size, which might be necessary for longer chains. {\color{red} For example, $\Delta h_\mathrm{max} = 1$ would be equivalent to allowing only (single-site) ``mutational" updates, i.e., picking randomly a position and changing the amino acid type at that position.}

\begin{figure}
\includegraphics[width=8.0cm]{Fig4}
\caption{Comparing sampling efficiencies of the MS and ST algorithms. Statistical errors $\sigma_{\left < E\right >}$ of the average total energy $\left < E\right >$ obtained for the sequences (A) A1, (B) N1, (C) R1 and (D) R2 (see Table~I) at different temperatures $T$. Simulation lengths in the two methods are adjusted such that the number of conformations sampled per sequence and temperature is roughly the same (see text). }
\end{figure}

@@

\begin{figure*}
\includegraphics[width=17.0cm]{Fig5}
\caption{Networks of sequences connecting folds IA and IB (top) and folds IIA and IIB (bottom). Each node represents a stable sequence ($\Ptot\ge\Pcut$ where $\Pcut=0.50$) that folds into either IA or IIA (light blue), IB or IIB (dark red), or is classified as bistable ($B>0.5$, black). A line between two nodes indicates that the sequences differ at only one position. Graph created using the tool Graphviz~\protect\cite{Graphviz2000} obtained from www.graphviz.org.}
\end{figure*}

\vspace{12pt}
\subsection{Exploring sequence space: IA/IB and IIA/IIB fold connectivities}

We now turn to the full binary sequence sets {\SI} and {\SII} with 1,024 sequences each. By applying the MS method to these two sets (see Table~II), we determine the low-$T$ thermodynamic behavior of each included sequence. In particular, we calculate the stabilities of folds IA and IB, $\PIA$ and $\PIB$, for all sequences in {\SI} and the stabilities of folds IIA and IIB, $\PIIA$ and $\PIIB$, for all sequences in {\SII} (see Methods). The relative statistical errors on these quantities vary but are only a few percent at the most, despite the large number of sequences included.
